Cleaning and Processing

### 2024-09-03 code/analysis-code/introanalysis-v1.qmd
Revisions for submission 5
clean up of code and text

```{r, message=FALSE,  warning=FALSE}
#load needed packages. make sure they are installed.
library(tidyverse) #for data processing/cleaning; includes ggplot2, tidyr, readr, dplyr, stringr, purr, forcats
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
library(knitr) #for nice tables
library(kableExtra) #for nice tables
library(gt) #for nice tables
library(gtsummary) #for summary tables
library(ggplot2) #for plotting
library(tidymodels) #for modeling
rm(list=ls()) #clear the environment
```

@@

After performing a Welch Two Sample t-test to compare the rates of LBW between Euro- and Afro-descent mothers, the results were as follows: t = `r round(t_statistic, 2)`, df = `r round(df, 1)`, p-value = `r round(p_value, digits = 3)`, mean difference = `r round(mean_diff, 4)`, 95% CI = [`r round(ci_lower, 4)`, `r round(ci_upper, 4)`].

@@

# Check the structure of the updated data frame
str(ML_linear)
```

@@

>>>>

```{r, message=FALSE,  warning=FALSE}
#load needed packages. make sure they are installed.
library(tidyverse) #for data processing/cleaning; includes ggplot2, tidyr, readr, dplyr, stringr, purr, forcats
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
library(knitr) #for nice tables
library(kableExtra) #for nice tables
library(gt) #for nice tables
library(gtsummary) #for summary tables
library(ggplot2) #for plotting
library(tidymodels) #for modeling
library(gtExtras) # for easy summary visualization
rm(list=ls()) #clear the environment
```

@@

After performing a Welch Two Sample t-test to compare the rates of LBW between Euro- and Afro-descent mothers, the results were as follows: t = `r round(t_statistic, 2)`, df = `r round(df, 1)`, p-value = `r round(p_value, digits = 3)`, mean difference = `r round(mean_diff, 4)`, 95% CI = \[`r round(ci_lower, 4)`, `r round(ci_upper, 4)`\].

@@

# Check the structure of the updated data frame
str(ML_linear)

# Save ML_Linear as RDS
save_data_location <- here::here("data","processed-data","ML_linear.rds")
saveRDS(ML_linear, file = save_data_location)
```

@@

# Predominant and residual distributions

```{r}
# Calculate the mean and standard deviation of the predominant distribution
predominant_stats <- ML_summary %>% 
   filter(!is.na(Weightgrams)) %>% #filter out NAs
  summarize(
    mean_weight = mean(Weightgrams),
    sd_weight = sd(Weightgrams)
  )

# Determine the range of the predominant distribution
predominant_lower <- predominant_stats$mean_weight - 2 * predominant_stats$sd_weight
predominant_upper <- predominant_stats$mean_weight + 2 * predominant_stats$sd_weight

# Calculate the proportion of observations in the predominant distribution
predominant_proportion <- ML_summary %>%
  filter(Weightgrams >= predominant_lower & Weightgrams <= predominant_upper) %>%
  nrow() / nrow(ML_summary)

# Calculate the proportion of observations in the residual distribution
residual_proportion <- 1 - predominant_proportion

# Print the results
cat("Predominant distribution:\n")
cat("Mean:", predominant_stats$mean_weight, "grams\n")
cat("Standard deviation:", predominant_stats$sd_weight, "grams\n")
cat("Proportion:", predominant_proportion, "\n\n")

cat("Residual distribution:\n")
cat("Proportion:", residual_proportion, "\n")

# Plot the distribution
residual_plot_all <- ML_summary %>% filter(!is.na(Weightgrams)) %>% #filter out NAs
  ggplot(aes(x = Weightgrams)) +
  geom_density(fill="#69b3a2", color="#e9ecef") +
  stat_function(fun = dnorm, args = list(mean = predominant_stats$mean_weight, sd = predominant_stats$sd_weight), color = "black", linewidth = 0.5) +
  labs(
    title = "Birth Weight Distribution for All Births, Maternidade Laranjeiras, 1922-1926",
    x = "Birth Weight (g)"
  ) +
  theme_ipsum() +
  theme(plot.title = element_text(size=12))
plot(residual_plot_all)

# Save plot to file
figure_file = here("results","figures","residual_plot_all.png")
ggsave(filename = figure_file, plot=residual_plot_all, bg="white") #background is white

# Calculate two distributions: Afro and Euro
# Calculate the mean and standard deviation of the predominant distribution Afro
predominant_stats_afro <- ML_summary %>% 
   filter(!is.na(WeightgramsAfro)) %>% #filter out NAs
  summarize(
    mean_weight_afro = mean(WeightgramsAfro),
    sd_weight_afro = sd(WeightgramsAfro)
  )

# Determine the range of the predominant distribution
predominant_lower_afro <- predominant_stats_afro$mean_weight_afro - 2 * predominant_stats_afro$sd_weight_afro
predominant_upper_afro <- predominant_stats_afro$mean_weight_afro + 2 * predominant_stats_afro$sd_weight_afro

# Calculate the proportion of observations in the predominant distribution
predominant_proportion_afro <- ML_summary %>%
  filter(WeightgramsAfro >= predominant_lower_afro & WeightgramsAfro <= predominant_upper_afro) %>%
  nrow() / nrow(ML_summary)

# Calculate the proportion of observations in the residual distribution
residual_proportion_afro <- 1 - predominant_proportion_afro

# Print the results
cat("Predominant distribution:\n")
cat("Mean:", predominant_stats_afro$mean_weight_afro, "grams\n")
cat("Standard deviation:", predominant_stats_afro$sd_weight_afro, "grams\n")
cat("Proportion:", predominant_proportion_afro, "\n\n")

cat("Residual distribution:\n")
cat("Proportion:", residual_proportion_afro, "\n")

# Calculate the mean and standard deviation of the predominant distribution Euro
predominant_stats_euro <- ML_summary %>% 
   filter(!is.na(WeightgramsEuro)) %>% #filter out NAs
  summarize(
    mean_weight_euro = mean(WeightgramsEuro),
    sd_weight_euro = sd(WeightgramsEuro)
  )

# Determine the range of the predominant distribution
predominant_lower_euro <- predominant_stats_euro$mean_weight_euro - 2 * predominant_stats_euro$sd_weight_euro
predominant_upper_euro <- predominant_stats_euro$mean_weight_euro + 2 * predominant_stats_euro$sd_weight_euro

# Calculate the proportion of observations in the predominant distribution
predominant_proportion_euro <- ML_summary %>%
  filter(WeightgramsEuro >= predominant_lower_euro & WeightgramsEuro <= predominant_upper_euro) %>%
  nrow() / nrow(ML_summary)

# Calculate the proportion of observations in the residual distribution
residual_proportion_euro <- 1 - predominant_proportion_euro

# Print the results
cat("Predominant distribution:\n")
cat("Mean:", predominant_stats_euro$mean_weight_euro, "grams\n")
cat("Standard deviation:", predominant_stats_euro$sd_weight_euro, "grams\n")
cat("Proportion:", predominant_proportion_euro, "\n\n")

cat("Residual distribution:\n")
cat("Proportion:", residual_proportion_euro, "\n")

# Plot both distributions
residual_plot_divided <- ML_summary %>% 
  filter(!is.na(WeightgramsAfro) | !is.na(WeightgramsEuro)) %>% #filter out NAs
  ggplot() +
  geom_density(aes(x = WeightgramsAfro), fill = "#69b3a2", alpha = 0.5, color = NA) +
  geom_density(aes(x = WeightgramsEuro), fill = "#404080", alpha = 0.5, color = NA) +
  stat_function(fun = dnorm, 
        args = list(mean = predominant_stats_afro$mean_weight_afro,
                sd = predominant_stats_afro$sd_weight_afro),
                color = "black", linewidth = 0.5) +
  stat_function(fun = dnorm, 
                args = list(mean = predominant_stats_euro$mean_weight_euro, 
                            sd = predominant_stats_euro$sd_weight_euro), 
                color = "black", linewidth = 0.5) +
  labs(
    title = "Birth Weight Distribution for Afro- and Euro-Descended Infants",
    subtitle = "Maternidade Laranjeiras, 1922-1926",
    x = "Birth Weight (g)",
    fill = "Infant Descent"
  ) +
  scale_x_continuous(limits = c(1000, 5000), breaks = seq(1000, 5000, by = 1000)) +
  theme_ipsum() +
  theme(plot.title = element_text(size=12),
        plot.subtitle = element_text(size=10),
        legend.position = "bottom")
plot(residual_plot_divided)

# Save plot to file
figure_file = here("results","figures","residual_plot_divided.png")
ggsave(filename = figure_file, plot=residual_plot_divided, bg="white") #background is white
```

### 2024-09-03 code/eda-code/eda-v1.qmd
Revisions for submission 5
clean up of code and text

```{r}
#Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_summary_processed.rds")

@@

Using these new dummy variables, I create new dummy variables defining if the infants were normal birth weight (NBW $>2500$ g) or low birth weight (LBW $\leq2500$ g) for all infants and then stratified by maternal skin color. These are `Afro_NBW` and `Afro_LBW`; and `Euro_NBW` and `Euro_LBW`. I will use these to see how many infants were low birth weight or normal birth weight for each maternal skin category in the `introanalysis-v1.qmd` file.

@@

# Create dummy variables for birth weight for infants born to Black, mixed-race, and White mothers for summary table. Create dummy variables for birth weight according to fetal sex.
ML_summary <- ML_summary %>%
  mutate(
    Weightblack = factor(if_else(Color == "Preta", "Yes", "No"), levels = c("No", "Yes")),
    Weightmixed = factor(if_else(Color == "Parda", "Yes", "No"), levels = c("No", "Yes")),
    Weightwhite = factor(if_else(Color == "Branca", "Yes", "No"), levels = c("No", "Yes")),
    Weightfemale = factor(if_else(Sex == "F", "Yes", "No"), levels = c("No", "Yes")),
    Weightmale = factor(if_else(Sex == "M", "Yes", "No"), levels = c("No", "Yes")),
    WeightgramsBlack = if_else(Weightblack == "Yes", Weightgrams, NA_real_),
    WeightgramsWhite = if_else(Weightwhite == "Yes", Weightgrams, NA_real_),
    WeightgramsMixed = if_else(Weightmixed == "Yes", Weightgrams, NA_real_),
    WeightgramsFemale = if_else(Weightfemale == "Yes", Weightgrams, NA_real_),
    WeightgramsMale = if_else(Weightmale == "Yes", Weightgrams, NA_real_)
  )

@@

# View the updated dataframe
print(head(ML_summary))
```
Let's also change the observations from Portuguese to English for the remaining observations to clean up things for the summary table.

@@

I have now learned that `gtsummary` is a very easy way to create summary statistics (Table 1). In the note section at the end of this document, I also include the code I worked on to create the tables manually to show my learning process.

@@

# Create gtsummary table for final analysis
table1 <- 
  ML_summary %>%
  mutate(
    WeightgramsBlack = if_else(Weightblack == "Yes", Weightgrams, NA_real_),
    WeightgramsWhite = if_else(Weightwhite == "Yes", Weightgrams, NA_real_),
    WeightgramsMixed = if_else(Weightmixed == "Yes", Weightgrams, NA_real_),
    WeightgramsFemale = if_else(Weightfemale == "Yes", Weightgrams, NA_real_),
    WeightgramsMale = if_else(Weightmale == "Yes", Weightgrams, NA_real_)
  ) %>%
  tbl_summary(
    include = c(Color, ModifiedColor, ModifiedStatus, Age, ModifiedNationality, Birth, MaternalOutcome, FetalOutcome, Sex, Lengthcentimeters, Weightgrams, WeightgramsBlack, WeightgramsMixed, WeightgramsWhite, WeightgramsFemale, WeightgramsMale, BirthweightCategory),
    missing = "no",
    label = list(
      Color ~ "Color (n = 2,695)",
      ModifiedColor ~ "Ancestry (n = 2,695)", 
      ModifiedStatus ~ "Parity (n = 2,836)", 
      Age ~ "Maternal Age (n = 2,783)", 
      ModifiedNationality ~ "Combined Nationality (n = 2,773)", 
      Birth ~ "Birth Outcome (n = 2,761)", 
      MaternalOutcome ~ "Maternal Outcome (n = 2,829)", 
      FetalOutcome ~ "Fetal Outcome (n = 2,666)", 
      Sex ~ "Sex (n = 2,534)",
      Lengthcentimeters ~ "Birth Length (cm) (n = 2,405)", 
      Weightgrams ~ "Birth Weight (g) (n = 2,384)",
      WeightgramsBlack ~ "Black (g) (n = 643)",
      WeightgramsMixed ~ "Mixed-Race (g) (n = 675)",
      WeightgramsWhite ~ "White (g) (n = 950)",
      WeightgramsFemale ~ "Female (g) (n = 1,074)",
      WeightgramsMale ~ "Male (g) (n = 1,270)",
      BirthweightCategory ~ "Birth Weight Category (n = 2,231)"
    ),
    statistic = list(
       all_continuous() ~ "{mean} ({p25}, {p75})",
      WeightgramsBlack ~ "{mean} ({p25}, {p75})",
      WeightgramsMixed ~ "{mean} ({p25}, {p75})",
      WeightgramsWhite ~ "{mean} ({p25}, {p75})",
      WeightgramsFemale ~ "{mean} ({p25}, {p75})",
      WeightgramsMale ~ "{mean} ({p25}, {p75})"
    ),
    digits = list(WeightgramsBlack ~ 0, WeightgramsMixed ~ 0, WeightgramsWhite ~ 0, WeightgramsFemale ~ 0, WeightgramsMale ~ 0)
  ) %>%
  #add_n() #%>%
  modify_table_body( #indent these three rows
    ~.x %>%
      mutate(label = case_when(
        variable == "WeightgramsBlack" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
        variable == "WeightgramsWhite" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
        variable == "WeightgramsMixed" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
        variable == "WeightgramsFemale" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
        variable == "WeightgramsMale" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
        TRUE ~ label
      ))
  )
table1

@@

```{r}
#Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","Mean_Birthweights_processed.rds")

@@

@@

```{r}
#Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_linear_processed.rds")
#load data
ML_linear <- readRDS(data_location)
```

@@

# Create dummy variables for color and change to English using recode factor from dplyr
ML_linear$ModifiedColor <- recode_factor(ML_linear$Color, "Preta" = "Afro-Descent", "Parda" = "Afro-Descent", "Branca" = "Euro-Descent")

@@

# Create dummy variables for nationality and change to English
ML_linear$ModifiedNationality <- recode_factor(ML_linear$Nationality, "Alema" = "European", "Argentina" = "Latin American", "Austriaca" = "European", "Brasileira" = "Brazilian", "Espanhola" = "European", "Francesa" = "European", "Italiana" = "European", "Paraguaya" = "Latin American", "Polaca" = "European", "Portuguesa" = "European", "Rumania" = "European", "Russa" = "European", "Suiça" = "European", "Siria" = "Middle Eastern", "Uruguaya" = "Latin American")

>>>>

```{r}
#Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_summary.rds")

@@

Using these new dummy variables, I created new dummy variables defining if the infants were normal birth weight (NBW $>2500$ g) or low birth weight (LBW $\leq2500$ g) for all infants and then stratified by maternal skin color. These are `Afro_NBW` and `Afro_LBW`; and `Euro_NBW` and `Euro_LBW`. I will use these to see how many infants were low birth weight or normal birth weight for each maternal skin category in the `introanalysis-v1.qmd` file.

@@

# Create dummy variables for birth weight for infants born to Black, mixed-race, and White mothers for summary table. Create dummy variables for birth weight according to fetal sex.
ML_summary <- ML_summary %>%
  mutate(
    Weightblack = factor(if_else(Color == "Preta", "Yes", "No"), levels = c("No", "Yes")),
    Weightmixed = factor(if_else(Color == "Parda", "Yes", "No"), levels = c("No", "Yes")),
    Weightwhite = factor(if_else(Color == "Branca", "Yes", "No"), levels = c("No", "Yes")),
    Weightfemale = factor(if_else(Sex == "F", "Yes", "No"), levels = c("No", "Yes")),
    Weightmale = factor(if_else(Sex == "M", "Yes", "No"), levels = c("No", "Yes")),
    WeightgramsBlack = if_else(Weightblack == "Yes", Weightgrams, NA_real_),
    WeightgramsWhite = if_else(Weightwhite == "Yes", Weightgrams, NA_real_),
    WeightgramsMixed = if_else(Weightmixed == "Yes", Weightgrams, NA_real_),
    WeightgramsFemale = if_else(Weightfemale == "Yes", Weightgrams, NA_real_),
    WeightgramsMale = if_else(Weightmale == "Yes", Weightgrams, NA_real_),
    WeightgramsAfro = if_else(ModifiedColor == "Afro-Descent", Weightgrams, NA_real_),
    WeightgramsEuro = if_else(ModifiedColor == "Euro-Descent", Weightgrams, NA_real_)
  )

@@

# View the updated dataframe
print(head(ML_summary))
```
I will now change the observations from Portuguese to English for the remaining observations to clean up things for the summary table.

@@

`Gtsummary` is a very easy way to create summary statistics (Table 1). In the note section at the end of this document, I also include the code I worked on to create the tables manually to show my learning process.

@@

# Create gtsummary table for final analysis
table1 <- 
  ML_summary %>%
  mutate(
    WeightgramsBlack = if_else(Weightblack == "Yes", Weightgrams, NA_real_),
    WeightgramsWhite = if_else(Weightwhite == "Yes", Weightgrams, NA_real_),
    WeightgramsMixed = if_else(Weightmixed == "Yes", Weightgrams, NA_real_),
    WeightgramsAfro = if_else(ModifiedColor == "Afro-Descent", Weightgrams, NA_real_),
    WeightgramsEuro = if_else(ModifiedColor == "Euro-Descent", Weightgrams, NA_real_),
    WeightgramsFemale = if_else(Weightfemale == "Yes", Weightgrams, NA_real_),
    WeightgramsMale = if_else(Weightmale == "Yes", Weightgrams, NA_real_)
  ) %>%
  tbl_summary(
    include = c(Color, ModifiedColor, ModifiedStatus, Age, ModifiedNationality, Birth, MaternalOutcome, FetalOutcome, Sex, Lengthcentimeters, Weightgrams, WeightgramsBlack, WeightgramsMixed, WeightgramsWhite, WeightgramsAfro, WeightgramsEuro, WeightgramsFemale, WeightgramsMale, BirthweightCategory),
    missing = "no",
    label = list(
      Color ~ "Color",
      ModifiedColor ~ "Ancestry", 
      ModifiedStatus ~ "Parity", 
      Age ~ "Maternal Age", 
      ModifiedNationality ~ "Combined Nationality", 
      Birth ~ "Birth Outcome", 
      MaternalOutcome ~ "Maternal Outcome", 
      FetalOutcome ~ "Fetal Outcome", 
      Sex ~ "Sex",
      Lengthcentimeters ~ "Birth Length (cm)", 
      Weightgrams ~ "Birth Weight (g)",
      WeightgramsBlack ~ "Black",
      WeightgramsMixed ~ "Mixed-Race",
      WeightgramsWhite ~ "White",
      WeightgramsAfro ~ "Afro-Descent",
      WeightgramsEuro ~ "Euro-Descent",
      WeightgramsFemale ~ "Female",
      WeightgramsMale ~ "Male",
      BirthweightCategory ~ "Birth Weight Category"
    ),
    statistic = list(
       all_continuous() ~ "{mean} ({p25}, {p75})",
      WeightgramsBlack ~ "{mean} ({p25}, {p75})",
      WeightgramsMixed ~ "{mean} ({p25}, {p75})",
      WeightgramsWhite ~ "{mean} ({p25}, {p75})",
      WeightgramsFemale ~ "{mean} ({p25}, {p75})",
      WeightgramsMale ~ "{mean} ({p25}, {p75})"
    ),
    digits = list(WeightgramsBlack ~ 0, WeightgramsMixed ~ 0, WeightgramsWhite ~ 0, WeightgramsFemale ~ 0, WeightgramsMale ~ 0)
  ) %>%
  add_n() %>%
  modify_table_body( #indent these three rows
    ~.x %>%
      mutate(label = case_when(
        variable == "WeightgramsBlack" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
        variable == "WeightgramsWhite" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
        variable == "WeightgramsMixed" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
        variable == "WeightgramsAfro" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
        variable == "WeightgramsEuro" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
        variable == "WeightgramsFemale" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
        variable == "WeightgramsMale" ~ paste0("\u00A0\u00A0\u00A0\u00A0", label),
        TRUE ~ label
      ))
  )
table1

@@

```{r}
#Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","Mean_Birthweights.rds")

@@

# ML Linear Data

@@

```{r}
#Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_linear.rds")
#load data
ML_linear <- readRDS(data_location)
```

@@

# Create dummy variables for color and change to English using recode factor from dplyr
ML_linear$ModifiedColor <- factor(ifelse(ML_linear$Color %in% c("Black", "Mixed Race"), "Afro-Descent", "Euro-Descent"))

@@

# Create dummy variables for nationality and change to English
ML_linear$ModifiedNationality <- factor(case_when(
  ML_linear$Nationality %in% c("German", "Austrian", "Spanish", "French", "Italian", "Polish", "Portuguese", "Romanian", "Russian", "Swiss") ~ "European",
  ML_linear$Nationality %in% c("Argentine", "Paraguayan", "Uruguayan") ~ "Latin American",
  ML_linear$Nationality == "Brazilian" ~ "Brazilian",
  ML_linear$Nationality == "Syrian" ~ "Middle Eastern"
))

### 2024-09-03 code/processing-code/processingfile-v1.qmd
Revisions for submission 5
clean up of code and text

```{r}
# Save ML_Linear as RDS
save_data_location <- here::here("data","processed-data","ML_linear_processed.rds")
saveRDS(ML_linear, file = save_data_location)

# Save ML_summary as RDS
save_data_location <- here::here("data","processed-data","ML_summary_processed.rds")
saveRDS(ML_summary, file = save_data_location)
```

>>>>

```{r}
# Save ML_Linear as RDS
save_data_location <- here::here("data","processed-data","ML_linear.rds")
saveRDS(ML_linear, file = save_data_location)

# Save ML_summary as RDS
save_data_location <- here::here("data","processed-data","ML_summary.rds")
saveRDS(ML_summary, file = save_data_location)
```
